Storm Surges

### 2014-07-21 RoughWork/manuscript.tex
Reorganize model configuration section so that it flows like a paper.

The curvilinear orthogonal numerical grid is divided into 398 by 898 by 40 grid cells, which results in an almost uniform horizontal resolution with grid spacing approximately 440 m by 500 m. The 40 vertical $z$-levels are stretched gradually in order to achieve higher resolution in the surface layer, with 1 m vertical grid spacing down to about 10 m in depth. Below 10m the grid is stretched gradually to a maximum grid spacing of 27 m at the lowest layer. At the bottom boundary, partial $z$-levels are utilized in order to limit large changes in bathymetry across grid cells \citep{madec2012nemo}.

@@

\subsection{Boundary conditions and sub-grid scale processes}
%lateral and bottom friction, open boundaries, eta surface
The model includes two open boundaries that connect to the Pacific Ocean, the western boundary of the Strait of Juan de Fuca as well as Johnstone Strait at the north, both of which are forced with eight tidal constituents, temperature and salinity climatologies, and the sea surface height anomaly. The details of these forcing conditions are described below. At coastal boundaries, the partial slip boundary condition, an approximation to no slip, is used. The partial slip boundary condition allows one to include the frictional effects of lateral boundaries without the restrictive resolution required to represent the lateral boundary layer under no slip conditions. A lateral eddy viscosity of 20 m $^2$ s $^{-1}$ parametrizes horizontal friction and a lateral eddy diffusivity of 20.5 m $^2$ s $^{-1}$ is used. At the ocean surface, meteorological conditions and fresh water input from rivers are included and described in more detail below. Bottom friction is represented by a quadratic law for the bottom momentum flux with drag coefficient $C_D = 5\times 10^{-3}$. Vertical turbulence and mixing is calculated through the $k-\epsilon$ configuration of the generic length scale (GLS) turbulence closure \citep{umlauf2003generic} with background vertical eddy viscosity and diffusivity set to $1\times10^{-4}$ and $1\times10^{-5}$ m$^2$ s$^{-1}$ respectively. Details on the NEMO implementation of the partial slip lateral boundary condition, quadratic bottom friction law, and GLS turbulence closure scheme are provided by \citet{madec2012nemo}.

In addition to the equations of motion, a prognostic equation for the sea surface height is solved at each time step. The inclusion the sea surface height equation requires a fairly restrictive time step due to the presence of high speed surface gravity waves. As such, the split-explicit time stepping algorithm is employed, where the free surface and barotropic equations are solved with a smaller time step than that used for the other variables. The model time step and barotropic time step are 10 s and 2 s respectively.

\subsubsection{Tidal forcing} 
The model was forced by eight tidal constituents (K$_1$ ,O$_1$, P$_1$, Q$_1$, M$_2$, K$_2$, N$_2$, S$_2$) at the Juan de Fuca and Johnstone Strait boundaries. Tidal heights and currents at grid points along the Juan de Fuca boundary were extracted from Webtide, an online web prediction model for the northeast Pacific Ocean, which is based on work by \citet{foreman2000webtide}. The Johnstone Strait boundary was forced with current and elevation tidal harmonics measured and calculated by \citet{thomson1980johnstone} for the major M$_2$ and K$_1$ constituents. Additionally, O$_1$ and S$_2$ elevation harmonics from their measurements were employed. The remaining constituents were extrapolated from Webtide.

\subsubsection{Temperature and salinity}
Temperature and salinity at the Juan de Fuca boundary were taken from a weekly climatology which was created from results from a model covering the Salish Sea and the west coast of Vancouver Island \citep{massonfine2012}.  Their results, originally on s-levels were interpolated onto z-levels and then onto the NEMO horizontal grid.  To prepare the climatology all years (1995-2008) were averaged and results, approximately every 15 days, were interpolated to a weekly climatology.

\subsubsection{Sea Surface Height}
Sea surface height at the mouth of Juan de Fuca was set using values from the Tofino tide gauge.  A monthly climatology was produced using daily averages from 2000-2010, binning them by month, averaging and setting the yearly mean to zero.  For the storm surge simulations, hourly variations in sea surface height were used.  These values are the Tofino tide gauge values, de-tided and with the zero reset as for the climatology. The storm surge simulations also used the sea surface height anomaly from Port Hardy, forced at the northern boundary in Johnstone Strait.

\subsubsection{Open Boundary Conditions}

The model is relaxed to the forced temperature and salinity over the 10 grid points (about 5km) closest to the open boundaries, using a flow relaxation scheme \citep{engedahl1995use}. %other references?
The tidal forcing and sea surface height was used in the barotropic velocity forcing which used the NEMO Flather scheme \citep{flather1994storm, madec2012nemo}.
The baroclinic velocities at the boundary were set to be equal to the values inside the boundary (zero-gradient boundary conditions).  This scheme is not part of core NEMO.  Zero gradient conditions were chosen because the baroclinic velocity at the mouth of Juan de Fuca is primarily estuarine and thus set by density variations between inside and outside the domain.

\subsection{River forcing}
River input provides a significant volume of freshwater to the Salish Sea and can influence stratification, circulation and primary productivity. However, most rivers in the domain are not gauged so parameterizations were required to represent river flow. \citet{morrison2011rivers} provides a method for estimating freshwater runoff in the Salish Sea region based on precipitation. Monthly runoff volumes for each watershed for each year from 1970 to 2012 were acquired from \citet{morrison2011rivers}, as well as monthly averages.

@@

The monthly flow from each river was input as a point source in the three grid points closest to the surface at the model point closest to the mouth of each river. Incoming water was assumed to be fresh and at surface temperature. A total of 150 rivers were parameterised by this method.

%we could show the location of all the rivers on the overall location map, or on a picture of our domain, since the rivers are on the edges of the domain so they won't block any bathymetry information

\subsection{Atmospheric forcing}
The ocean surface is forced with momentum and heat fluxes from a 33-km global atmospheric reforcasting model suitable for use in ocean modelling \citep{smith2013new}. Forecasts from the period of 2002-2011 are available. Additionally, the inverse barometric effect of the atmospheric pressure is included, an important consideration in storm surge modelling.

\subsection{Initial conditions}
Initial conditions for temperature and salinity were taken from a CTD cast in the middle Strait of Georgia taken in Sept 2002 \citep{pawlowiczetal2007}.  Conditions were initially uniform horizontally.  Velocity was initialized at zero. 
%Merge with below?
%Perhaps image of spun-up stratification along Thalweg? Here or in model evaluation

\subsection{Spin-up}
The model was spun up for a 15.5 months from the initial conditions above, starting Sep 16, 2002, using atmospheric forcing from 2002-2003, climatological temperature and salinity and sea surface height at the boundaries, with tides and climatological river output.  All storm surge runs were started three days prior to the event of interest with zero initial velocities and sea surface height and a stratification profile from model spin up. The modelled sea surface height adjusted to forcing in less than one day.

@@

\begin{equation}
D = [(A_0 \cos g_0 - A_m \cos g_m)^2 + (A_0 \sin g_0 - A_m \sin g_m)^2]^{1/2}
\end{equation}
where $A_0$, $A_m$, $g_0$ and $g_m$ are observed and modelled amplitudes and phases.

@@

\begin{table}[h]
\centering 
\begin{tabular}{l l c c c c c c} 
\hline 
& & \multicolumn{3}{c}{M$_2$} & \multicolumn{3}{c}{K$_1$} \\ 
\hline 
Station number & Station name & $R$ & $ \Delta \phi$ (deg) & $D$ (cm) & $R$ & $ \Delta \phi$ (deg) & $D$ (cm) \\
\hline 
               &              &     &                      &          &     &                      &          \\
\hline   
Mean Error     &              &     &                      &          &     &                      &          \\
RMS Error      &              &     &                      &          &     &                      &          \\
\hline  
\end{tabular}
\caption{Comparisons of modelled and observed amplitudes and phases. $R$ is the ration of modelled to observed amplitude. $\Delta \phi$ is the difference between modelled and observed phase. $D$ is the complex difference. }
\label{tab:comparison} 
\end{table}
%May have to include station numbers and locations on a map? Might be too busy with rivers too...
%Should we use the new method of calculating tides?

@@

\subsection{Stratification} 
%outline: thalweg, plots of winter/summer stratty
%Maybe note this

@@

Point Atkinson observed a maximum surge amplitude of 84 cm, which is approximately 20 cm higer than the model result, however the timing of the surge agrees almost exactly. The best agreement occurs at Victoria where the observed surge was 62 cm and the model returned 54 cm. At all of the locations, the model predicted the maximum surge to occur within 3.5 hours of the observed surge. Typically, the modelled surge was too early, in contrast to the later predictions in the Feb, 4 2006 case. Errors in the modelled water level at the timing of maximum surge were of the same order as the errors in the surge amplitude.

>>>>

The curvilinear orthogonal numerical grid has been divided into 398 by 898 by 40 grid cells, which results in an almost uniform horizontal resolution with grid spacing approximately 440 m by 500 m. The 40 vertical $z$-levels were stretched gradually in order to achieve higher resolution in the surface layer, with 1 m vertical grid spacing down to about 10 m in depth. Below 10m the grid was stretched to a maximum grid spacing of 27 m at the lowest layer. At the bottom boundary, partial $z$-levels were utilized in order to limit large changes in bathymetry across grid cells \citep{madec2012nemo}.

In addition to the equations of motion, a prognostic equation for the sea surface height is solved at each time step. The inclusion of the sea surface height equation requires a fairly restrictive time step due to the presence of high speed surface gravity waves. As such, the split-explicit time stepping algorithm is employed, where the free surface and barotropic equations are solved with a smaller time step than that used for the other variables. The model time step is and barotropic time step are 10 s and 2 s respectively. Small vertical grid spacing and large vertical velocities in the Boundary Pass and Haro Strait region impose a fairly restricitive step through the vertical CFL condition.

@@

\subsection{Boundary conditions and sub-grid scale processes}
The model includes two open boundaries that connect to the Pacific Ocean, the western boundary of the Strait of Juan de Fuca as well as Johnstone Strait at the north, both of which are forced with eight tidal constituents (K$_1$ ,O$_1$, P$_1$, Q$_1$, M$_2$, K$_2$, N$_2$, S$_2$), temperature and salinity climatologies, and the sea surface height anomaly. Tidal heights and currents at grid points along the Juan de Fuca boundary were extracted from Webtide, an online web prediction model for the northeast Pacific Ocean, which is based on work by \citet{foreman2000webtide}. The Johnstone Strait boundary was forced with current and elevation tidal harmonics measured and calculated by \citet{thomson1980johnstone} for the major M$_2$ and K$_1$ constituents. Additionally, O$_1$ and S$_2$ elevation harmonics from their measurements were employed. The remaining constituents were extrapolated from Webtide. 
% Needs editing! Update when we are finished with tides.

Temperature and salinity at the Juan de Fuca boundary were taken from a weekly climatology which was created from results from a model covering the Salish Sea and the west coast of Vancouver Island \citep{massonfine2012}.  Their results, originally on s-levels were interpolated onto z-levels and then onto the NEMO horizontal grid.  To prepare the climatology all years (1995-2008) were averaged and results, approximately every 15 days, were interpolated to a weekly climatology. The model is relaxed to the forced temperature and salinity over the 10 grid points (about 5km) closest to the open boundaries, using a flow relaxation scheme \citep{engedahl1995use}. %other references?

Sea surface height at the mouth of Juan de Fuca was set using values from the Tofino tide gauge.  A monthly climatology was produced using daily averages from 2000-2010, binning them by month, averaging and setting the yearly mean to zero.  For the storm surge simulations, hourly variations in sea surface height were used.  These values are the Tofino tide gauge values, de-tided and with the zero reset as for the climatology. The storm surge simulations also used the sea surface height anomaly from Port Hardy, forced at the northern boundary in Johnstone Strait. The tidal forcing and sea surface height was used in the barotropic velocity forcing which used the NEMO Flather scheme \citep{flather1994storm, madec2012nemo}.
The baroclinic velocities at the boundary were set to be equal to the values inside the boundary (zero-gradient boundary conditions).  This scheme is not part of core NEMO.  Zero gradient conditions were chosen because the baroclinic velocity at the mouth of Juan de Fuca is primarily estuarine and thus set by density variations between inside and outside the domain.

At coastal boundaries, the partial slip boundary condition, an approximation to no slip, is used. Partial slip allows one to include the frictional effects of lateral boundaries without the restrictive resolution required to represent the lateral boundary layer under no slip conditions. A lateral eddy viscosity of 20 m $^2$ s $^{-1}$ parametrizes horizontal friction and a lateral eddy diffusivity of 20.5 m $^2$ s $^{-1}$ is used.  Bottom friction is represented by a quadratic law for the bottom momentum flux with drag coefficient $C_D = 5\times 10^{-3}$. Vertical turbulence and mixing is calculated through the $k-\epsilon$ configuration of the generic length scale (GLS) turbulence closure \citep{umlauf2003generic} with background vertical eddy viscosity and diffusivity set to $1\times10^{-4}$ and $1\times10^{-5}$ m$^2$ s$^{-1}$ respectively. Details on the NEMO implementation of the partial slip lateral boundary condition, quadratic bottom friction law, and GLS turbulence closure scheme are provided by \citet{madec2012nemo}.

At the ocean surface, meteorological conditions and fresh water input from rivers are included. The ocean surface is forced with momentum and heat fluxes from a 33-km global atmospheric reforcasting model suitable for use in ocean modelling \citep{smith2013new}. Forecasts from the period of 2002-2011 are available. Additionally, the inverse barometric effect of the atmospheric pressure is included, an important consideration in storm surge modelling.

River input provides a significant volume of freshwater to the Salish Sea and can influence stratification, circulation and primary productivity. However, most rivers in the domain are not gauged so parameterizations were required to represent river flow. \citet{morrison2011rivers} provides a method for estimating freshwater runoff in the Salish Sea region based on precipitation. Monthly runoff volumes for each watershed for each year from 1970 to 2012 were acquired from \citet{morrison2011rivers}, as well as monthly averages.

@@

The monthly flow from each river was input as a point source in the three grid points closest to the surface at the model point closest to the mouth of each river. Incoming water was assumed to be fresh and at surface temperature. A total of 150 rivers were parameterised by this method and their locations are indicated by the blue dots in Figure \ref{fig:domain}.

Initial conditions for temperature and salinity were taken from a CTD cast in the middle Strait of Georgia taken in Sept 2002 \citep{pawlowiczetal2007}.  Conditions were initially uniform horizontally and velocity was initialized at zero. The model was spun up for a 15.5 months from the initial conditions above, starting Sep 16, 2002, using atmospheric forcing from 2002-2003, climatological temperature and salinity and sea surface height at the boundaries, with tides and climatological river output.  All storm surge runs were started three days prior to the event of interest with zero initial velocities and sea surface height and a stratification profile from model spin up. The modelled sea surface height adjusted to forcing in less than one day.

@@

\begin{equation}
D = [(A_0 \cos g_0 - A_m \cos g_m)^2 + (A_0 \sin g_0 - A_m \sin g_m)^2]^{1/2}
\end{equation}\label{eq:compdiff}
where $A_0$, $A_m$, $g_0$ and $g_m$ are observed and modelled amplitudes and phases.

@@

\begin{table}[h]
\centering 
\begin{tabular}{l l c c c c c c} 
\hline 
& & \multicolumn{3}{c}{M$_2$} & \multicolumn{3}{c}{K$_1$} \\ 
\hline 
Station number & Station name & $R$ & $ \Delta \phi$ (deg) & $D$ (cm) & $R$ & $ \Delta \phi$ (deg) & $D$ (cm) \\
\hline 
               &              &     &                      &          &     &                      &          \\
\hline   
Mean Error     &              &     &                      &          &     &                      &          \\
RMS Error      &              &     &                      &          &     &                      &          \\
\hline  
\end{tabular}
\caption{Comparisons of modelled and observed amplitudes and phases. $R$ is the ratio of modelled to observed amplitude. $\Delta \phi$ is the difference between modelled and observed phase. $D$ is the complex difference. }
\label{tab:comparison} 
\end{table}
%May have to include station numbers and locations on a map? Maybe on the contour map, overlay with stations and green/red dots (hopefully mostly green)
%Should we use the new method of calculating tides?

@@

\subsection{Stratification} 
%outline: thalweg, plots of winter/summer stratty
%Maybe not this

@@

Point Atkinson observed a maximum surge amplitude of 84 cm, which is approximately 20 cm higer than the model result, however the timing of the surge agrees almost exactly. The best agreement occurs at Victoria where the observed surge was 62 cm and the model returned 54 cm. At all of the locations, the model predicted the maximum surge to occur within 3.5 hours of the observed surge. Typically, the modelled surge was too early, in contrast to the later predictions in the Feb, 4 2006 case. Errors in the modelled water level at the time of maximum surge were of the same order as the errors in the surge amplitude.
